Generalized Algebraic Data Types in Haskell

### 2013-05-27 gadt.tex
Comments
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

This article is a tutorial about one such extension -- generalized algebraic data types (GADTs). The single idea of GADTs is to allow specific return types of value constructors. In this way they generalize ordinary algebraic data types. The theoretical foundation for GADTs is the notion of dependent types which is extensively described in literature on computer science and logic~\cite{pierce_book}.

@@

The key feature of GADTs is that pattern matching causes type refinement. In the right-hand side of the following equation the type of \verb|a| is refined to \verb|Int|:

@@

Examples in the following sections show the real practical value of GADTs.

@@

This section introduces GADTs with a canonical example of expression evaluator. At first, we attempt to implement it using ordinary algebraic data types. But as we will see, GADTs allow a more elegant implementation of the evaluator.

@@

Evaluation function for such expressions is also tricky. The result of evaluation can be either an integer or a Boolean value. The type \verb|ExtExpr| is not parametrized by return value type, so we have to use type \verb|Either Int Bool|. Also, evaluation will fail if the input expression is incorrect, so we have to use type \verb|Maybe|. Finally, type signature is the following:

@@

And implementation of this function is complicated. For example, processing \verb|AddInt| requires usage of a nested \verb|case|:

@@

Ideally we want the following type signature for \verb|evaluate| method:

@@

Using GADT syntax the data type \verb|PhantomExpr| type can be declared this way:

@@

All value constructors have \verb|PhantomExpr t| as their return type. As noted in the previous section, the distinctive feature of GADTs is the ability to return specific types in value constructors, for example \verb|PhantomExpr Int|. GADT for the expression language looks this way:

@@

Pattern matching causes type refinement, so for example in the right-hand side of the following expression \verb|i| has type \verb|Int|:

@@

At the end of this article we describe one more implementation of expression evaluator.

@@

In datatype-generic programming functions take a type as argument and their behavior depends on the structure of this type. There are several approaches to such kind of generic programming in Haskell. Paper by Hinze et al.~\cite{generic_programming} provides an overview of these approaches. In this section we present an approach which uses GADTs. Ideas for this section were taken from another paper by Hinze at al.~\cite{fun_with_phantom_types}.

Suppose we would like to write a function to encode data in binary form. This function must be able to work with values of several types. Functions like this one can be implemented using type classes. But GADTs offer an interesting alternative.

@@

\begin{Verbatim}
ghci> let d = [Dyn TInt 10, Dyn tString "test"]
ghci> :t d
d :: [Dynamic]
\end{Verbatim}

However, we can't make this list a \verb|Dynamic| value itself. To fix this problem, we need to extend the representation type: add a value constructor for the \verb|Dynamic| data type.

@@

The \verb|Dynamic| data type is useful for communication with the environment when the actual type of the data is not known in advance. In this case it is required to use a type cast to get the useful data. A simple way to implement a type cast from \verb|Dynamic| data type to an integer is the following:

@@

There is a more generic solution~\cite{fun_with_phantom_types} of this problem that works for all types, not just integer, but it is out of scope for this article.

@@

An important role of type systems is to ensure that data is manipulated in appropriate ways (for example, to make sure that we pass a list to \verb|head| function). But types can be used to express more sophisticated properties. For example, we can define a type of lists of a particular length and then define \verb|headSafe| function that only accepts non-empty lists. The idea described in this section is to use types to express correctness properties and then use type checker of the programming language to ensure that we can express only those programs that have the desired properties. This idea was developed in $\Omega$mega system~\cite{curry_howard,omega}. In this section we GADTs in Haskell programming language to prove correctness of list operations and the next section describes proving correctness of insertion operation in red-black trees.

@@

Actually, with the current data type definition a term \verb|Cons 'a' Nil| can even be given type \verb|SafeList Char Int|. To fix this problem it is required to give the same kind for \verb|Empty| and \verb|NonEmpty| types. This is discussed later for \verb|Nat| data type.

But \verb|SafeList| data type does not have enough static information to prove list length invariants for list functions. For example, for the concatenation function we need to show that length of the concatenated list is a sum of source lists lengths. So it is not enough to just know if a list is empty or not. We need to encode the length of a list in its type.

@@

Zero is encoded as \verb|Zero|, one -- as \verb|Succ Zero|, two -- as \verb|Succ (Succ Zero)| and so on. Now list data type is defined as:

@@

To implement the concatenation function we need a type-level function for addition of Peano numbers. A natural way to implement such function is to use \italic{type families} (which in context of this tutorial can be understood as type-level functions). First we need to declare type family \verb|Plus| (this requires \verb|TypeFamilies| extension):

@@

Then we need to declare type instances that implement addition of Peano numbers by induction:

@@

In fact, at the moment \verb|Succ| has type parameter of kind \verb|*|, so it possible to write nonsensical terms like \verb|Succ Int|, and they will be accepted by the type checker. This problem can be addressed using a new \italic{kind}. Just as types classify values, kinds classify types. We can declare the following data type:

@@

Here \verb|Nat| is a type; \verb|Zero| and \verb|Succ| are value constructors. But due to promotion~\cite{promotion} \verb|Nat| also becomes a kind; \verb|Zero| and \verb|Succ| also become types (see \prettyref{fig:promotion}). Where necessary, a quote must be used to resolve ambiguity. For example, \verb|'Succ| refers to a type, not a value constructor. So, type-level representation of Peano number 2 can be written as:

@@

On the one hand, the count parameter must be passed as a value to populate the list at run-time. On the other hand, we need a type-level representation of the same number for \verb|List| type. Haskell enforces a phase separation between run-time values and compile-time types.

@@

In a function returning an element by index in the list we need to make sure that the index does not exceed the list length. This requires a type-level function to compute whether one number is less than the other. We define the following type family and instances (\verb|TypeOperators| extension is required to be able to define \verb|:<| operation for types):

@@

@@

@@

Now we see that color must be passed as a type (for \verb|IncBlackHeight| type family) and as a value (to the value constructor). Similarly as before, we need to use a singleton type as a bridge:

@@

Because normally when creating a new type in Haskell, every type variable that appears on the right-hand side of the definition must also appear on its left-hand side. One solution to this problem is usage of \italic{existential types} (this definition requires extension \verb|RankNTypes|):

@@

It is also possible to do this with GADT:

@@

Proving the 3rd invariant is more involved. First we need to specify valid colors for a node on the type level. This can be done using type families as before or using type classes. We choose the latter and define a type class with 3 parameters corresponding to color of the parent and colors of the child nodes (this code requires \verb|MultiParamTypeClasses| extension):

@@

The implementation of the insertion operation can temporarily invalidate the 3rd invariant (see \prettyref{fig:balancing}), so during insertion we are not able to represent the tree using this data type. Thus it is required to declare a data type similar to \verb|Node|, but without the restriction on node colors:

@@

As before we need to make changes in type annotations of the functions implementing insert operation. We also need to change the \verb|leftBalance| function type signature this way:

@@

Earlier we passed the whole node as a parameter. But we can't do it after the \verb|Node| data type was modified: the 3rd invariant could be violated due to insertion in the left branch of the node. So, we pass all parameters of the parameters of the node and left child is represented using \verb|IntNode| data type.

@@

@@

Hindley-Milner (HM) is a classical type inference method~\cite{cardelli_typechecking}. One of the most important properties of HM is ability to always deduce the \italic{most general type} (\italic{principle type}) of every term.

@@

Yampa~\cite{yampa} is a domain-specific language for functional reactive programming (FRP). FRP is a programming paradigm of expressing data flows using the building blocks of functional programming. Based on the information from the paper by Nilsson~\cite{yampa_gadts}, this section describes how GADTs were used to improve performance of Yampa programs.

Signal function is a central abstraction in Yampa. It represents a simple synchronous process that maps an input signal to an output signal (see \prettyref{fig:signal_function}). The type of the signal function is \verb|SF a b| and it can be constructed from an ordinary function using the following function:

@@

There are other performance improvements that are enabled by GADTs in Yampa~\cite{yampa_gadts}. The results of performance improvements are shown in the table. It was taken from the paper by Nilsson~\cite{yampa_gadts}. The table shows execution time of several benchmarks using initial simply-optimized implementation ($T_S$) and the implementation with GADT-based optimizations ($T_G$). In addition to performance improvements, GADTs allowed to write a more concise and cleaner API without the need of pre-composed signal function (that were defined in Yampa only for performance reasons).

@@

@@

>>>>

This article is a tutorial about one such extension -- generalized algebraic data types (GADTs). The single idea of GADTs is to allow specific return types of value constructors. \EZY{I think this explanation doesn't mean anything to someone who doesn't already know what GADTs are.} In this way they generalize ordinary algebraic data types. The theoretical foundation for GADTs is the notion of dependent types which is extensively described in literature on computer science and logic~\cite{pierce_book}.

@@

\EZY{I think I'd add something like, ``While creating a values with the TI results in a value of type Test Int, the converse holds too: when a pattern-match succeeds on the TI constructor, we can now conclude that a is Int; that is to say, pattern matching causes type refinement.''}
The key feature of GADTs is that pattern matching causes type refinement. In the right-hand side of the following equation the type of \verb|a| is refined to \verb|Int|:

@@

Examples in the following sections show some practical applications of GADTs.

@@

This section introduces GADTs with the canonical example of an expression evaluator. At first, we attempt to implement it using ordinary algebraic data types. But as we will see, GADTs allow a more elegant implementation of the evaluator.

@@

Evaluation function for such expressions is also tricky. The result of evaluation can be either an integer or a Boolean value. The type \verb|ExtExpr| is not parametrized by return value type, so we have to use type \verb|Either Int Bool|. Furthermore, evaluation will fail if the input expression is incorrect, so we have to use type \verb|Maybe|. Finally, the type signature of \verb|evaluate| is the following:

@@

Implementing this function is complicated. For example, processing \verb|AddInt| requires usage of a nested \verb|case|:

@@

Ideally, we want the following type signature for \verb|evaluate| method:

@@

Using GADT syntax, the data type \verb|PhantomExpr| type can be declared this way:

@@

All value constructors have \verb|PhantomExpr t| as their return type. As noted in the previous section, the distinctive feature of GADTs is the ability to return specific types in value constructors, for example \verb|PhantomExpr Int|. The expression language can be written with GADTs in this fashion:

@@

Pattern matching causes type refinement: the right-hand side of the following expression \verb|i| has type \verb|Int|:

@@

At the end of this article, we describe one more implementation of expression evaluator.

@@

In datatype-generic programming, functions take a type as an argument and change behavior depending on the structure of this type. There are several approaches to such kind of generic programming in Haskell. A paper by Hinze et al.~\cite{generic_programming} provides an overview of these approaches. In this section, we present an approach which uses GADTs. Ideas for this section were taken from another paper by Hinze at al.~\cite{fun_with_phantom_types}.

Suppose we would like to write a function to encode data in binary form. This function must be able to work with values of several types. Functions like this one can be implemented using type classes. However, GADTs offer an interesting alternative.

@@

\begin{Verbatim}
ghci> let d = [Dyn TInt 10, Dyn TString "test"]
ghci> :t d
d :: [Dynamic]
\end{Verbatim}

However, we can't make this list a \verb|Dynamic| value itself. To fix this problem, we need to extend the representation type, adding a value constructor for the \verb|Dynamic| data type.

@@

The \verb|Dynamic| data type is useful for communication with the environment when the actual type of the data is not known in advance. In this case a type cast is required to get useful data. A simple way to implement a type cast from \verb|Dynamic| data type to an integer is the following:

@@

There is a more generic solution~\cite{fun_with_phantom_types} for this problem that works for all types, not just integer, but it is out of scope for this article.

@@

An important role of type systems is to ensure that data is manipulated in appropriate ways (for example, to ensure we pass a list to \verb|head| function). But types can be used to express more sophisticated properties. For example, we can define a type of lists of a particular length and then define a \verb|headSafe| function that only accepts non-empty lists. The idea described in this section is to use types to express correctness properties and then use type checker of the programming language to ensure that we can express only those programs that have the desired properties. This idea was developed in $\Omega$mega system~\cite{curry_howard,omega}. In this section, we use GADTs in the Haskell programming language to prove the correctness of list operations; in the next section, we describe how to use GADTs to prove correctness of the insertion operation in red-black trees.

@@

It's worth noting that with the current data type definition, a term \verb|Cons 'a' Nil| can even be given the type \verb|SafeList Char Int|. To fix this problem, we need to somehow give the \verb|Empty| and \verb|NonEmpty| types the same kind. This is discussed later for \verb|Nat| data type.

Regardless, \verb|SafeList| data type does not have enough static information to prove stronger list length invariants for many list functions. For example, for the concatenation function we need to show that length of the concatenated list is a sum of source lists lengths. It is not enough to just know if a list is empty or not: we need to encode the length of a list in its type.

@@

Zero is encoded as \verb|Zero|, one -- as \verb|Succ Zero|, two -- as \verb|Succ (Succ Zero)| and so on. Now the list data type is defined as:

@@

To implement the concatenation function, we need a type-level function for addition of Peano numbers. A natural way to implement such function is to use \italic{type families} (which in context of this tutorial can be understood as type-level functions). First, we need to declare type family \verb|Plus| (this requires \verb|TypeFamilies| extension):

@@

Then, we need to declare type instances that implement addition of Peano numbers by induction:

@@

As mentioned previously, \verb|Succ| has a type parameter of kind \verb|*|, so it possible to write nonsensical terms like \verb|Succ Int| which will be accepted by the type checker. This problem can be addressed using a new \italic{kind}. Just as types classify values, kinds classify types. We can declare the following data type:

@@

Here \verb|Nat| is a type; \verb|Zero| and \verb|Succ| are value constructors. But due to promotion~\cite{promotion} \verb|Nat| also becomes a kind; \verb|Zero| and \verb|Succ| also become types (see \prettyref{fig:promotion}). Where necessary, a quote must be used to resolve ambiguity. For example, \verb|'Succ| refers to a type, not a value constructor. So, type-level representation of the number two can be written as:

@@

On the one hand, the count parameter must be passed as a value to populate the list at run-time. On the other hand, we need a type-level representation of the same number for \verb|List| type. Haskell enforces a phase separation between run-time values and compile-time types \EZY{so there is a problem. Please state what the problem is.}.

@@

In a function returning an element by index in the list we need to make sure that the index does not exceed the list length. This requires a type-level function to compute whether one number is less than the other. We define the following type family and instances (the \verb|TypeOperators| extension is required to be able to define \verb|:<| operation for types):

@@

\EZY{Yes, please write a conclusion here!}

@@

\EZY{Before you dive into all of the hoary details of the construction, it is well worth motivating the example. What's the point? Answer this question!}

@@

Now we see that a color must be passed as a type (for \verb|IncBlackHeight| type family) and as a value (to the value constructor). Similarly as before, we need to use a singleton type as a bridge:

@@

since normally when creating a new type in Haskell, every type variable that appears on the right-hand side of the definition must also appear on its left-hand side. One solution to this problem is usage of \italic{existential types} (this definition requires extension \verb|RankNTypes|):

@@

It is also possible to do this with a GADT:

@@

Proving the 3rd invariant is more involved. First, we need to specify valid colors for a node on the type level. This can be done using type families as before or using type classes. We choose the latter and define a type class with 3 parameters corresponding to color of the parent and colors of the child nodes (this code requires \verb|MultiParamTypeClasses| extension):

@@

The implementation of the insertion operation can temporarily invalidate the 3rd invariant (see \prettyref{fig:balancing}), so during insertion we are not able to represent the tree using this data type. Thus, we need to declare a data type similar to \verb|Node|, but without the restriction on node colors:

@@

As before, we need to make changes in type annotations of the functions implementing insert operation. We also need to change the \verb|leftBalance| function type signature this way:

@@

Earlier, we passed the whole node as a parameter. But we can't do this after the \verb|Node| data type was modified: the 3rd invariant could be violated due to insertion in the left branch of the node. So, we pass all parameters of the parameters of the node and left child is represented using \verb|IntNode| data type.

@@

\EZY{Conclusion???}

@@

\EZY{This section is something of a digression from the litany of examples that have been parading down this article; so it's probably worth pointing it out and mentioning why this is an important thing to know about.}

Hindley-Milner (HM) is the classic type inference method~\cite{cardelli_typechecking}. One of the most important properties of HM is ability to always deduce the \italic{most general type} (\italic{principle type}) of every term.

@@

Yampa~\cite{yampa} is a domain-specific language for functional reactive programming (FRP). FRP is a programming paradigm which involves expressing data flows using the building blocks of functional programming. \EZY{What are the building blocks of functional programming?} Based on the information from the paper by Nilsson~\cite{yampa_gadts}, this section describes how GADTs were used to improve performance of Yampa programs.

The signal function is a central abstraction in Yampa. It represents a simple synchronous process that maps an input signal to an output signal (see \prettyref{fig:signal_function}). The type of the signal function is \verb|SF a b| and it can be constructed from an ordinary function using the following function:

@@

There are other performance improvements that are enabled by GADTs in Yampa~\cite{yampa_gadts}. The results of performance improvements described by the paper by Nilsson~\cite{yampa_gadts} are shown in the table. The table shows execution time of several benchmarks using initial simply-optimized implementation ($T_S$) and the implementation with GADT-based optimizations ($T_G$). In addition to performance improvements, GADTs allowed the authors of the paper to write a more concise and cleaner API without the need of pre-composed signal functions (which were defined in Yampa only for performance reasons).

@@

\EZY{What are the limitations of finally tagless in this context?}

@@

\EZY{I think you can write a better concluding sentence then that!}
